Measuring contrast processing in the visual system using SSVEP

### 2024-05-06 review.qmd
More BCI

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006].

Electrophysiological and psychophysical measurements of the effect of attention on both luminance and chromatic contrast have strongly implicated gain control as an underlying mechanism [@DiRusso1999; @DiRusso2001; @Lu1998]. These effects also appear to differ between chromatic and achromatic pathways [@DiRusso1999b] - perhaps as a result of the different levels of nonlinear gain control in the early pre-cortical magno-, parvo- and konio-cellular pathways [@Derrington1984b; @Kaplan1986; @Lee1990; @Solomon2005]. In the late 2000s a comprehensive theoretical model for attentional modulation was developed that framed it as a gain control computation [@Reynolds2009; @Boynton2009]. This framework has proven to be influential - explaining a wide range of phenomena from the earlier literature and demonstrating subtle interactions between the size of the attentional 'spotlight' and the stimulus configuration which rationalise many apparent contradictions in the literature. Direct measurements of attentional modulation of achromatic SSVEP signals are broadly consistent with this model [@Lauritzen2010, @Hou2016, @Martinovic2018] and confirm the relatively weaker role of spatial attention on responses driven by chromatic stimuli - particularly those that isolate the opponent S-(L+M) cone pathway [@Wang2011, @Highsmith2010]. The gain control model of attention can also be extended to SSVEP studies of feature-based attention which show that the modulatory effects can be targeted to the most informative neuronal populations [@Verghese2012].

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated). EEG signals are highly sensitive to both visual field position [@Ales2010] and attentional state [@Verghese2012; @Lauritzen2010], so the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm.

Probably a summary of BCIs here @Zhu2010

>>>>

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006, @Muller1998m @MullerHillyard2000, MorganPNAS1996].

Electrophysiological and psychophysical measurements of the effect of attention on both luminance and chromatic contrast have strongly implicated gain control as an underlying mechanism [@DiRusso1999; @DiRusso2001; @Lu1998]. These effects also appear to differ between chromatic and achromatic pathways [@DiRusso1999b] - perhaps as a result of the different levels of nonlinear gain control in the early precotical magno-,  parvo- and konio-cellular pathways [DerringtonLennie84, KaplanShapley86, LeePokornyJOSA90, SolomonLennie2005].  In the late 2000s a comprehensive theoretical model for attentional modulation was developed that framed it as a gain control computation [@Reynolds2009; @Boynton2009]. This framework has proven to be influential - explaining a wide range of phenomena from the earlier literature and demonstrating subtle interactions between the size of the attentional 'spotlight' and the stimulus configuration which rationalised many apparent contradictions in the literature. Direct measurements of attentional modulation of contrast-driven SSVEP signals are broadly consistent with this model [@Lauritzen2010, @Hou2016, @Martinovic2018] and confirm the relatively weaker role of spatial attention on responses driven by chromatic stimuli - particularly those that isolate the opponent S-(L+M) cone pathway [@Wang2011, @Highsmith2010]. The gain control model of attention can also be extended to SSVEP studies of feature-based attention which show that the modulatory effects can be targeted selectively to the most informative neuronal populations [@Verghese2012,@Ding2006].

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [Middendorf2000, ]. EEG signals are highly sensitive to both visual field position [@Ales2010] and attentional state [@Verghese2012; @Lauritzen2010], so the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties : For example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli to increase SNR or the range of stimuli that can be frequency-tagged (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....).
